### 2015-08-05 background.tex
Removed vspace hacks for now.
Done up to Section 3

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver. The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter. However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break; // racy
    ...
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in a Linux 4.0 driver (drivers/char/generic\_nvram.c)}
\label{fig:data_race_example}
\vspace{-2mm}
\end{figure}

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break;
    ...
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\vspace{-2mm}
\end{figure}

\noindent\textbf{Lockset Analysis }
%
Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.  The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location. This is because an empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, $\mathit{CLS}_T$ is empty for every $T$, because the threads do not hold any locks on program start. The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs), because a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. This is because the tool runs the program in a truly concurrent environment, thus it only gets to see what the thread did on the particular schedule that was followed.

To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis in a static context, using dataflow analysis. In this paper, we present a novel symbolic lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\vspace{-5mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
\vspace{-3mm}
\end{figure}

>>>>

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux 4.0 driver. The driver can invoke the entry point concurrently from two threads, with the same \texttt{file} struct as a parameter. This can lead to multiple possible data races because the threads can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break; // racy
    ...
  }
  if (offset < 0) return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux 4.0 driver}
\label{fig:data_race_example}
%\vspace{-2mm}
\end{figure}

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable \texttt{res} inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by regarding them as not providing any protection between threads; this can lead to false alarms.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    ...
    case 1: offset += file->f_pos; break;
    ...
  }
  if (offset < 0) {
    mutex_unlock(&nvram_mutex); // unlock
    return -EINVAL;
  }
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Introducing a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
%\vspace{-2mm}
\end{figure}

\noindent\textbf{Lockset Analysis }
%
Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.  The idea is to track the set of locks (i.e.\ \emph{lockset}) that are \emph{consistently} used to protect a memory location during program execution. An empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads. Consequently, the analysis reports a \emph{potential} race on that memory location.

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, every $\mathit{CLS}_T$ is empty because the threads do not hold any locks on program start. In addition, every $\mathit{LS}_s$ is initialized to the set of all locks manipulated by the program since initially each access to $s$ is (vacuously) protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

In contrast to more sophisticated race analyses that encode a \emph{happens-before} relation between threads~\cite{lamport1978time} (e.g.\ using vector clocks), lockset analysis is easy to implement, lightweight, and thus has the potential to scale well.  The technique, though, suffers from imprecision (i.e.\ can report false bugs) since a violation of the locking discipline does not always correspond to a real data race~\cite{savage1997eraser, pozniansky2003efficient, o2003hybrid, elmas2007goldilocks, flanagan2009fasttrack}. Furthermore, the code coverage in dynamic lockset analyzers, such as Eraser, is limited by the execution paths that are explored under a given scheduler. Since the tool runs the program in a truly concurrent environment, it only gets to see what the thread did on the particular schedule that was followed.

To counter the above limitations, techniques such as Locksmith~\cite{pratikakis2006locksmith} and RELAY~\cite{voung2007relay} have explored the idea of applying lockset analysis statically, using dataflow analysis. In this paper, we present a novel symbolic lockset analysis method that involves generating verification conditions, which are then discharged to a theorem prover.

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
%\vspace{-5mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
%\vspace{-3mm}
\end{figure}

### 2015-08-05 intro.tex
Removed vspace hacks for now.
Done up to Section 3

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} found that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were found to be \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

@@

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel.  By combining \whoop and \corral we achieve analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as concurrent systems that use a similar programming model (e.g.\ file systems).

To summarize, our contributions are as follows:
\vspace{-0.5mm}
\begin{itemize}
\item We propose symbolic pairwise lockset analysis, a sound and scalable technique for automatically verifying the absence of data races in device drivers.
\vspace{-0.2mm}
\item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
\vspace{-0.2mm}
\item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, an industrial-strength bug-finder.
\vspace{-0.2mm}
\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race-checking and accelerating \corral.
\end{itemize}

>>>>

Device drivers are complex pieces of system-level software, operating at the thin boundary between hardware and software to provide an interface between the operating system and hardware devices that are attached to a computer. Drivers are notoriously hard to develop and debug~\cite{corbet2005linux}. This has a negative impact on hardware product releases, as time-to-market is commonly dominated by driver development, verification, and validation schedules~\cite{yavatkar2012era}.
%
Even after a driver has shipped, it typically has many undetected errors: Chou et al.~\cite{chou2001empirical} gathered data from 7 years of Linux kernel releases and found that the relative error-rate in driver source code is up to 10 times higher than in any other part of the kernel, while Swift et al.~\cite{Swift2003windowsxp} found that 85\% of the system crashes in Windows XP are due to faulty drivers. Regarding \emph{concurrency bugs}, a recent study~\cite{ryzhyk2009dingo} established that they account for 19\% of the total bugs in Linux drivers, showcasing their significance. The majority of these concurrency bugs were \emph{data races} or \emph{deadlocks} in various configuration functions and hot-plugging handlers.

@@

We present \whoop, an automated approach for static data race analysis in device drivers. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, which attempts to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are consistently protected by a common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We show that we can exploit the race-freedom guarantees provided by our symbolic analysis to achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a precise bug-finder used by Microsoft to analyze drivers and other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the Linux 4.0 kernel.  Our combination achieves analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observed even greater speedups of 12$\times$ and 20$\times$.
%
\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems, as well as concurrent systems that use a similar programming model (e.g.\ file systems).

To summarize, our contributions are as follows:
%\vspace{-0.5mm}
\begin{itemize}
\item We propose symbolic pairwise lockset analysis, a sound and scalable technique for automatically verifying the absence of data races in device drivers.
%\vspace{-0.2mm}
\item We present \whoop, a tool that leverages our approach for analyzing data races in device drivers.
%\vspace{-0.2mm}
\item We show that we can achieve a sound partial-order reduction using our technique to accelerate \corral, an industrial-strength bug-finder.
%\vspace{-0.2mm}
\item We analyze \sizeOfBenchmarks Linux drivers, showing that \whoop is efficient at race-checking and accelerating \corral.
\end{itemize}
